Lexicalized Domain Representations for Neural Machine Translation

### 2019-08-27 2019/iwslt/submission.tex
additional domain scenario

\section{Introduction \label{sec:introduction}}
Owing to the development of flexible and powerful architectures based on neural networks \cite{Cho14properties,Bahdanau15learning,Ghering17convolutional,Vaswani17attention}, Machine Translation (MT) has made significant progresses over the past years, and constitute to date the standard for most production engines. The development of MT systems, be they neural or statistical, require very large parallel corpora consisting of millions of sentence pairs, a resource that only exist in very few application domains and language pairs. A lot of the recent research effort has thus focused on developping MT systems in restricted data conditions, for instance building multilingual MTs which enable zero-shot translation \cite{Firat16multiway,Ha16towards,Johnson17google};
Another important scenario for industrial MT is to adapt a neural system trained using parallel data in one domain to the peculiarity of other domains. Domain Adaption (DA) in MT is an old issue \cite{Foster07mixture,Axelrod11domain}, which comes in various guises, and for which a number of solutions have been studied. See the recent survey of \cite{Chu18asurvey} for neural MT. The typical setting is \emph{supervised adaptation}, where a (small) amount of data of the target domain of interest is used to fine-tune the parameters of a system trained on a large amount of texts in a source domain. We study here a different scenario, \emph{multi-domain adaptation} \cite{Sennrich13multidomain,Farajian17multidomain}, where we would like to use heterogenous data sources to train a unique system that would work well for all domains. This allows us to be both data efficient (all data is used to train all domains) and computationally efficient (we only train one system).
Multi-domain adaptation is conceptually close to multilingual MT, or more generally to multi-task learning \cite{Caruana97multitask}, and can be approached in a number of ways. We adapt here ideas of \cite{Daume07frustratingly} to neural MT.
Our main hypothesis is that domains mostly differ at the lexical level, due to cross-domain polysemy, which motivates domain specific embeddings. By contrast, the deeper layers, which arguably model more abstract linguistic phenomena, are made shareable across domains. To this end, we design word embeddings containing a generic and several domain-specific regions. We experiment with three domains, two neural architectures and two language pairs and find that our technique yields effective multi-domain NMTs, outperforming several baselines. Our contribution is thus as follows:
we adapt and implement the ideas of \cite{Daume07frustratingly} for two NMT architectures;
we provide experimental evidence that show the effectiveness of this technique;
\jcTodo{highlight the analysis of embeddings as an original technique.}
our analysis of embeddings comforts our assumption that their variation across domains actually reflects a variation of senses.
\jcTodo{introduce the additional domain scenario experiment.}

@@

We experiment with two language pairs (English-French, English-German) and data originated from three domains, corresponding to texts from three European institutions: 
the European Parlement (EPPS) \cite{Koehn05europarl}, 
the European Medicines Agency (EMEA), 
the European Central Bank (ECB) \cite{Tiedemann09news},
In addition, for English-French we also used IT-domain corpora obtained from the OPUS web site\footnote{\url{http://opus.nlpl.eu}} corresponding to KDE, Ubuntu, GNOME and PHP datasets (IT).

We randomly split those corpora into training, validation and test sets (see statistics in Table~\ref{tab:Corpora}).
Validation sets are used to chose the best model according to the average BLEU score \cite{Papineni02bleu}.
Using internal test sets corresponds to somewhat ideal conditions.

@@

\jcTodo{I delete vocabs since they make no sense if computed after BPE}

@@

%The right parts of Tables~\ref{tab:results-trsf} and \ref{tab:results-rnn} report performance on test sets for new domains not seen in training. 
%Khresmoi texts, which are extracted from summaries of medical articles are assumed to be from the same domain as EMEA data; 
%for the other two test sets, we assume no domain and only use the generic part of the embeddings. 
%Here, we see that our best system \texttt{LDR}$_{oracle}$ is slightly lagging behind the \texttt{Mixed} baselines (the difference ranges from unsignificant to one Bleu point), which can be attributed to larger useful (generic) embeddings in the \texttt{Mixed} setting. 
%Here again, we see that using the sole generic part makes the system robust with respect to unknown or misleading domain tags. 
%We finally outline that our approach outperforms DC for these test sets (were we have to commit to a predicted domain tag), and even outperforms WDCMT (for German-English). 
%This might be due to the fact that WDCMT has no built-in mechanism to ignore, as we do in the \texttt{LDR}$_{generic}$ systems, the possibly misleading predicted domain information.
\jcTodo{compute averages}

@@

\jcTodo{introduce WDCMT experiments}
Finally the multi-domain model of \cite{Zeng18multidomain} (\texttt{WDCMT}).
The comparison with the \texttt{WDCMT} model is less favourable for our RNN systems, which are arguably much simpler.
The direct comparison with WDCMT is difficult as the implementation differs in many ways from our RNN: a different framework, different cell types,  etc.
  % (LSTMs vs. conditional GRUs),

\begin{table}[!h]
\begin{center}
\scalebox{1.0}{
\begin{tabular}{|l|ccc|c|}
\hline
Model & EMEA & EPPS & ECB & Avg. \\
\hline
%\multicolumn{8}{l}{} \\[-9pt]
\multicolumn{5}{l}{English$\rightarrow$French} \\
\hline
$\mathtt{WDCMT}$           & 68.76 $_{45.29}$ & 35.71 & 52.75 & \\
$\mathtt{LDR_{condgru}}_{pred}$   & 71.70 $_{46.21}$ & 35.09 & 51.22 & \\
\hline
\end{tabular}
} %scalebox
\end{center}
\caption{BLEU scores for RNN systems\label{tab:results-rnn-wdcmt}}
\end{table}

@@

\begin{table}[!h]
\begin{center}
\scalebox{1.0}{
\begin{tabular}{|l|ccc|c|}
\hline
$LDR$ & EMEA & EPPS & ECB & Avg. \\
\hline
\multicolumn{5}{l}{English$\rightarrow$French} \\
\hline
 size=4   & 74.65 $_{49.61}$ & 37.42 & 54.49 & \\
 size=8   & 74.26 $_{49.90}$ & 37.67 & 54.07 & \\
 size=16 & 74.15 $_{49.10}$ & 37.78 & 54.56 & \\
 size=32 & 75.10 $_{48.61}$ & 37.64 & 54.29 & \\
 size=64 & 74.50 $_{50.17}$ & 37.27 & 54.50 & \\
\hline
\end{tabular}
} %scalebox
\end{center}
\caption{BLEU scores for the Transformer architecture for varying domain-specific embedding sizes \label{tab:embedding-size}}
\end{table}

@@

We evaluate now the ability of our model to cope with new unseen domains, a very common scenario for industrial MT.
We simulate the apparition of the IT domain (see Section~\ref{ssec:data}) once the $\mathtt{LDR}_{oracle}$ model is built for EMEA, EPPS and ECB domains.
Note that to accommodate a new domain the network must reserve the new IT region from the beginning.
Hence, word embeddings contain 32 cells (8 x 4 domain regions) and 480 cells for the generic region.
Equivalently to our previous $\mathtt{LDR}_{oracle}$ model, 200,000 iterations are performed using EMEA, EPPS and ECB data sets (the IT region is never updated).
Then, 100,000 additional iterations are performed using  EMEA, EPPS, ECB and IT data sets ($\mathtt{LDR}_{oracle}^*$).
Experiments are only reported for the English$\rightarrow$French language pair.
For comparison purposes we also continue training our $\mathtt{Mixed}$ model applying the same number of iterations and using the same four datasets ($\mathtt{Mixed}^*$).
BLEU results are shown in Table~\ref{tab:add}.

\begin{table}[!h]
\begin{center}
\scalebox{0.9}{
\begin{tabular}{|l|cccc|c|}
\hline
Model & EMEA & EPPS & ECB & IT & Avg.\\
\hline
\multicolumn{5}{l}{English$\rightarrow$French} \\
\hline
$\mathtt{Mixed}$                     & 67.69$_{47.60}$ & 37.50 & 53.49 & 13.91 & \\
$\mathtt{Mixed}^*$             & 66.49$_{45.79}$ & 37.59 & 55.07 & 51.78 & \\
\hline
$\mathtt{LDR}_{oracle}$          & 74.26$_{49.90}$ & 37.67 & 54.07 & 13.40 & \\
$\mathtt{LDR}_{oracle}^*$  & 76.17$_{49.71}$ & 37.48 & 55.12 & 55.24 & \\
\hline
\end{tabular}
} %scalebox
\end{center}
\caption{BLEU scores for the Transformer architecture for varying domain-specific embedding sizes \label{tab:add}}
\end{table}

As expected, a huge gap is observed over the IT test sets when using in-domain data.
Note that $\mathtt{LDR}_{oracle}^*$ clearly outperforms $\mathtt{Mixed}^*$ by a large margin.
As previously, all models show very little differences on the EPPS test set.
Both models show similar results over the ECB test set when further trained with IT data.
Only our $\mathtt{LDR}_{oracle}$ model shows gains on the EMEA test set when further trained (only on the large test set).
Overall better results for our LDR when further trained.
\jcTodo{rewrite this part}

>>>>

\section{Introduction \label{sec:introduction}}
Owing to the development of flexible and powerful architectures based on neural networks \cite{Cho14properties,Bahdanau15learning,Ghering17convolutional,Vaswani17attention}, 
Machine Translation (MT) has made significant progresses over the past years, and constitute to date the standard for most production engines. 
The development of MT systems, be they neural or statistical, require very large parallel corpora consisting of millions of sentence pairs, a resource that only exist in very few application domains and language pairs. 
A lot of the recent research effort has thus focused on developing MT systems in restricted data conditions, for instance building multilingual MTs which enable zero-shot translation \cite{Firat16multiway,Ha16towards,Johnson17google};
Another important scenario for industrial MT is to adapt a neural system trained using parallel data in one domain to the peculiarity of other domains.

Domain Adaption (DA) in MT is an old issue \cite{Foster07mixture,Axelrod11domain}, which comes in various guises, and for which a number of solutions have been studied. 
See the recent survey of \cite{Chu18asurvey} for neural MT. 
The typical setting is \emph{supervised adaptation}, where a (small) amount of data of the target domain of interest is used to fine-tune the parameters of a system trained on a large amount of texts in a source domain. 
We study here a different scenario, \emph{multi-domain adaptation} \cite{Sennrich13multidomain,Farajian17multidomain}, where we would like to use heterogenous data sources to train a unique system that would work well for all domains. 
This allows us to be both data efficient (all data is used to train all domains) and computationally efficient (we only train one system).
Multi-domain adaptation is conceptually close to multilingual MT, or more generally to multi-task learning \cite{Caruana97multitask}, and can be approached in a number of ways.

We adapt here ideas of \cite{Daume07frustratingly} to neural MT.
Our main hypothesis is that domains mostly differ at the lexical level, due to cross-domain polysemy, which motivates domain specific embeddings. 
By contrast, the deeper layers, which arguably model more abstract linguistic phenomena, are made shareable across domains. 
To this end, we design word embeddings containing a generic and several domain-specific regions. 
We experiment with three domains, two neural architectures and two language pairs and find that our technique yields effective multi-domain NMTs, outperforming several baselines. 
Our contribution is thus as follows:
we adapt and implement the ideas of \cite{Daume07frustratingly} for two NMT architectures;
we provide experimental evidence that show the effectiveness of this technique;
we evaluate the ability of our network to efficiently accommodate new domains;
and we introduce a novel technique to analyse word polysemy using word embeddings which comforts our assumption that their variation across domains actually reflects a variation of senses.

@@

We experiment with two language pairs (English-French, English-German) and data originated from three domains, corresponding to texts from three European institutions: 
the European Parlement (EPPS) \cite{Koehn05europarl}, 
the European Medicines Agency (EMEA), 
the European Central Bank (ECB) \cite{Tiedemann09news},
In addition, for English-French we also used IT-domain corpora obtained from the OPUS web site\footnote{\url{http://opus.nlpl.eu}} corresponding to KDE, Ubuntu, GNOME and PHP datasets (IT).
%
We randomly split those corpora into training, validation and test sets (see statistics in Table~\ref{tab:Corpora}).
Validation sets are used to chose the best model according to the average BLEU score \cite{Papineni02bleu}.
%Using internal test sets corresponds to somewhat ideal conditions.

@@

@@

%The right parts of Tables~\ref{tab:results-trsf} and \ref{tab:results-rnn} report performance on test sets for new domains not seen in training. 
%Khresmoi texts, which are extracted from summaries of medical articles are assumed to be from the same domain as EMEA data; 
%for the other two test sets, we assume no domain and only use the generic part of the embeddings. 
%Here, we see that our best system \texttt{LDR}$_{oracle}$ is slightly lagging behind the \texttt{Mixed} baselines (the difference ranges from unsignificant to one Bleu point), which can be attributed to larger useful (generic) embeddings in the \texttt{Mixed} setting. 
%Here again, we see that using the sole generic part makes the system robust with respect to unknown or misleading domain tags. 
%We finally outline that our approach outperforms DC for these test sets (were we have to commit to a predicted domain tag), and even outperforms WDCMT (for German-English). 
%This might be due to the fact that WDCMT has no built-in mechanism to ignore, as we do in the \texttt{LDR}$_{generic}$ systems, the possibly misleading predicted domain information.
\jcTodo{compute averages on all tables}

@@

We also compare our presented network with the multi-domain model of \cite{Zeng18multidomain} (\texttt{WDCMT}) for English$\rightarrow$French translation.
\jcTodo{(explain why conditional GRU are used for the comparison)},
Results are shown by Table~\ref{tab:results-rnn-wdcmt}. 
The comparison is less favourable for our RNN systems, which are arguably much simpler.\footnote{The direct comparison with \texttt{WDCMT} is difficult as the implementation differs in many ways from our RNN: framework, cell types,  etc.}

\begin{table}[!h]
\begin{center}
\scalebox{1.0}{
\begin{tabular}{|l|ccc|c|}
\hline
Model & EMEA & EPPS & ECB & Avg. \\
\hline
%\multicolumn{8}{l}{} \\[-9pt]
\multicolumn{5}{l}{English$\rightarrow$French} \\
\hline
$\mathtt{WDCMT}$           & 68.76 $_{45.29}$ & 35.71 & 52.75 & \\
$\mathtt{LDR_{condgru}}_{pred}$   & 71.70 $_{46.21}$ & 35.09 & 51.22 & \\
\hline
\end{tabular}
} %scalebox
\end{center}
\caption{BLEU scores when comparing \texttt{LDR} to \texttt{WDCMT}.\label{tab:results-rnn-wdcmt}}
\end{table}

@@

\begin{table}[!h]
\begin{center}
\scalebox{1.0}{
\begin{tabular}{|l|ccc|c|}
\hline
$\mathtt{LDR}_{oracle}$ & EMEA & EPPS & ECB & Avg. \\
\hline
\multicolumn{5}{l}{English$\rightarrow$French} \\
\hline
 size=4   & 74.65 $_{49.61}$ & 37.42 & 54.49 & \\
 size=8   & 74.26 $_{49.90}$ & 37.67 & 54.07 & \\
 size=16 & 74.15 $_{49.10}$ & 37.78 & 54.56 & \\
 size=32 & 75.10 $_{48.61}$ & 37.64 & 54.29 & \\
 size=64 & 74.50 $_{50.17}$ & 37.27 & 54.50 & \\
\hline
\end{tabular}
} %scalebox
\end{center}
\caption{BLEU scores for the Transformer architecture for varying domain-specific embedding sizes \label{tab:embedding-size}}
\end{table}

@@

We evaluate now the ability of our model to integrate new domains, a very common scenario for industrial MT.
We simulate the apparition of the IT domain once the $\mathtt{LDR}_{oracle}$ model is built for EMEA, EPPS and ECB domains.
Note that to accommodate a new domain the network must be designed with a region reserved for the new domain.
Hence, for this experiment word embeddings contain 480 cells for the generic region and 32 cells for domain specific regions (8 cells x 4 regions).
Equivalently to our previous $\mathtt{LDR}_{oracle}$ model, 200,000 iterations are performed using EMEA, EPPS and ECB data sets (the IT region is never updated).
Then, 100,000 additional iterations are performed using  EMEA, EPPS, ECB and IT datasets ($\mathtt{LDR}_{oracle}^*$).
Experiments are reported for the English$\rightarrow$French language pair.
For comparison purposes we also continue training our $\mathtt{Mixed}$ model applying the same number of iterations and using the same four datasets ($\mathtt{Mixed}^*$).
BLEU results are shown in Table~\ref{tab:add}.

\begin{table}[!h]
\begin{center}
\scalebox{0.9}{
\begin{tabular}{|l|cccc|c|}
\hline
Model & EMEA & EPPS & ECB & IT & Avg.\\
\hline
\multicolumn{5}{l}{English$\rightarrow$French} \\
\hline
$\mathtt{Mixed}$                & 67.69$_{47.60}$ & 37.50 & 53.49 & 13.91 & \\
$\mathtt{Mixed}^*$             & 66.49$_{45.79}$ & 37.59 & 55.07 & 51.78 & \\
\hline
$\mathtt{LDR}_{oracle}$     & 74.26$_{49.90}$ & 37.67 & 54.07 & 13.40 & \\
$\mathtt{LDR}_{oracle}^*$  & 76.17$_{49.71}$ & 37.48 & 55.12 & 55.24 & \\
\hline
\end{tabular}
} %scalebox
\end{center}
\caption{BLEU scores for the Transformer architecture when including IT as additional domain \label{tab:add}}
\end{table}

As expected, a huge gap is observed over the IT test set when learning includes in-domain data.
Note that $\mathtt{LDR}_{oracle}^*$ clearly outperforms $\mathtt{Mixed}^*$ by a large margin over the IT test set.
%
Considering the other test sets: both models similarly enhance their performance on the ECB test set;
$\mathtt{LDR}_{oracle}^*$ shows gains over the EMEA test set while a loss of quality is observed on $\mathtt{Mixed}^*$;
using IT data does not impact the quality of translations of the EPPS test set of any of the models.
Overall better results are obtained by our $\mathtt{LDR}_{oracle}$ model when trained including the additional domain dataset.
New domains are easily integrated in the model if the corresponding embedding regions are reserved in advance.
